DynaProg for Scala

### 2013-01-04 docs/report/implementation.tex
More report.

\newpage
\section{Implementation}
% ------------------------------------------------------------------------------------------------
\subsection{CUDA ad-hoc}
In the project planning, an ad-hoc implementation phase immediately followed the problem analysis (we also present the parallelogram matrix case). The goal of this phase is threefold:\ol
\item Better understand the challenges in CUDA implementation of dynamic programming problems and get on par with state-of-art implementations.
\item Have a baseline implementation that is independent of the hardware and that could be benchmarked. We also tried to contact the authors of \cite{swat_mega} and \cite{gpu_atlp} to obtain their implementation. The former provided us with their implementation, which turned out to address large serial problems whereas our focus was on smaller non-serial problems, the latter did not respond to our solicitations.
\item Have an optimal implementation that can serve as target to be imitated and generalized by the code generation.
\ole

@@

% ------------------------------------------------------------------------------------------------
\subsection{Scala parsers}
The Scala parsers consist in 4 traits that are used to construct a DSL program:\ul
\item \textbf{Signature:} abstraction to define input and output types, and the aggregation function. The signature is implemented by all other traits (in particular algebras and grammars).
\item \textbf{BaseParsers:} serves as basis for the two other traits and defines common features. It implements the {\tt Parser} abstraction and all its inheriting classes: {\tt Tabulate}, (abstract) {\tt Terminal}, {\tt Aggregate}, {\tt Filter}, {\tt Map}, {\tt Or}, {\tt Concat}. Terminals are further specialized in the two other traits (ADPParsers and TTParsers). The parser abstraction specifies 3 methods:\ul
	\item {\tt apply(subword)} computes the parser result; it is used to obtain the corresponding results.
	\item {\tt unapply(subword,backtrack)} computes the previous step of the backtrack by returning subsequences at the origin of the result; it is invoked recursively to obtain the full backtrack trace.
	\item {\tt reapply(subword,backtrack)} is very similar to apply, except that it  computes only the results matching the backtrack. It is used to construct the result corresponding to a backtrack trace (possibly in a different domain, pretty printing, ...).
	\ule
	Additionally, the BaseParser implements the analysis that are shared by both the Scala and the CUDA version: dead rules elimination, yield analysis and dependencies ordering. Finally, it provides some implicit functions to flatten nested tuples (that are constructed by multiple concatenations).
\item \textbf{ADPParsers:} used as basis for a single track DP grammar (using one input sequence). It defines the concatenation operator $\sim$ ({\tt Concat} wrapper), and the terminals (empty, element and sequence). Additionally, it defines the interface functions {\tt parse(input)}, {\tt backtrack(input)} and {\tt build(in,backtrack)} that respectively compute the result, the backtrack and the result corresponding to a trace.
\item \textbf{TTParsers:} used to define two-track DP grammar (using a pair of sequences as input). Similarly, this class defines concatenations $-\!\!\sim$ and $\sim\!\!-$, terminals (for each track) and the {\tt parse(in1,in2)}, {\tt backtrack(in1,in2)} and {\tt build(in1,in2,backtrack)} functions.
\ule

@@

XXX: make a class diagram ? (classes hierarchies)

XXX: explain thoroughly all the functionalities and dig into the code ?

\subsection{Transformations and simplifications}

@@

@@

\subsection{Code generation}

\subsection{Runtime execution engine}

\subsection{CUDA implementation}

>>>>

\newpage
\section{Implementation}
% ------------------------------------------------------------------------------------------------
\subsection{Ad-hoc CUDA}
In the project planning, an ad-hoc implementation phase immediately followed the problem analysis (we also present the parallelogram matrix case). The goal of this phase is threefold:\ol
\item Better understand the challenges in CUDA implementation of dynamic programming problems and get on par with state-of-art implementations.
\item Have a baseline implementation that is independent of the hardware and that could be benchmarked. We also tried to contact the authors of \cite{swat_mega} and \cite{gpu_atlp} to obtain their implementation. The former provided us with their implementation, which turned out to address large serial problems whereas our focus was on smaller non-serial problems, the latter did not respond to our solicitations.
\item Have an optimal implementation that can serve as target to be imitated and generalized by the code generation.
\ole

@@

% ------------------------------------------------------------------------------------------------
\subsection{Scala parsers}
The Scala parsers consist in 4 traits that are used to construct a DSL program:\ul
\item \textbf{Signature:} abstraction to define input ({\tt Alphabet}) and output ({\tt Answer}) types, and the aggregation function. The signature is implemented by all other traits (in particular algebras and grammars).
\item \textbf{BaseParsers:} serves as basis for the two other traits and defines common features. It implements the {\tt Parser} abstraction and all its inheriting classes: {\tt Tabulate}, (abstract) {\tt Terminal}, {\tt Aggregate}, {\tt Filter}, {\tt Map}, {\tt Or}, {\tt Concat}. Terminals are further specialized in the two other traits (ADPParsers and TTParsers). The parser abstraction specifies 3 methods:\ul
	\item {\tt apply(subword)} computes the parser result; it is used to obtain the corresponding results.
	\item {\tt unapply(subword,backtrack)} computes the previous step of the backtrack by returning subsequences at the origin of the result; it is invoked recursively to obtain the full backtrack trace.
	\item {\tt reapply(subword,backtrack)} is very similar to apply, except that it  computes only the results matching the backtrack. It is used to construct the result corresponding to a backtrack trace (possibly in a different domain, pretty printing, ...).
	\ule
	To support analysis, the parsers carry additional values:\ul
	\item Minimum and maximum yield size: functions evaluated recursively except for tabulations where value is attributed in the yield analysis phase.
	\item Number of inner alternatives: helps counting alternatives, hereby guaranteeing an unique number for each (provided that parsers obtain non-overlapping ranges).
	\item Number of inner moving concatenations: helps determining required storage for the backtrack as well as retrieving the appropriate index in the backtrack phase
	\ule
	Additionally, the BaseParser implements the analysis that are shared by both the Scala and the CUDA version: dead rules elimination, yield analysis and dependencies ordering. Finally, it provides some implicit functions to flatten nested tuples (that are constructed by multiple concatenations).
\item \textbf{ADPParsers:} used as basis for a single track DP grammar (using one input sequence). It defines the concatenation operator $\sim$ ({\tt Concat} wrapper), and the terminals (empty, element and sequence). Additionally, it defines the interface functions {\tt parse(input)}, {\tt backtrack(input)} and {\tt build(in,backtrack)} that respectively compute the result, the backtrack and the result corresponding to a trace.
\item \textbf{TTParsers:} used to define two-track DP grammar (using a pair of sequences as input). Similarly, this class defines concatenations $-\!\!\sim$ and $\sim\!\!-$, terminals (for each track) and the {\tt parse(in1,in2)}, {\tt backtrack(in1,in2)} and {\tt build(in1,in2,backtrack)} functions.
\ule

\begin{figure}[H]\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(14,11)
\put(3,8){\tbox{8}{2.5}{{\bf Signature} \footnotesize\\ Types: Alphabet, Answer \\ $h$ (aggregation function)}}
\put(0,4){\tbox{14}{2.5}{{\bf BaseParsers} \footnotesize\\ Tabulate, Terminal, Aggregate, Filter, Map, Or, Concat \\ Analysis: dead rules, yield analysis, dependencies}}
\put(0,0){\tbox{6}{2.5}{{\bf ADPParsers} \footnotesize\\ $\sim$, $\sim(a,b,c,d)\sim$ \\ Single track terminals}}
\put(8,0){\tbox{6}{2.5}{{\bf TTParsers} \footnotesize\\ $-\!\!\sim$, $\sim\!\!-$ \\ Two-tracks terminals}}
{\linethickness{1.5pt}\put(3,2.5){\vector(1,1){1.5}}\put(11,2.5){\vector(-1,1){1.5}}\put(7,6.5){\vector(0,1){1.5}}}
\end{picture}\end{center}\caption{Class diagram (simplified)}\end{figure}

% ------------------------------------------------------------------------------------------------
\newpage
\subsection{Code generation}

@@

@@

\subsection{JNI wrappers and runtime execution engine}

@@

### 2013-01-04 docs/report/report.tex
More report.

\subsection*{Abstract}
Dynamic programming is a common pattern of Computer Science used in various domains. Yet underlying matrix recurrences might be difficult to express and error prone. Additionally, domain experts might not have the skills to make an efficient parallel implementation. In this project, we leverage the Scala and LMS infrastructure to implement a domain specific language (DSL) for dynamic programming on heterogeneous platforms, which allows to write concise and efficient programs. Our contributions are: \ul
\item A classification of DP problems characteristics (matrix shape, dependency graph, ...)
\item A systematic approach to process data (top-down/bottom-up) and backtracking information (focus on running time and memory efficiency)
\item A language embedded in Scala (DSL) to express DP problems concisely (based on ADP)
\item Two implementations: Scala for CPU (features) and an CUDA for GPU (efficiency)
%\item Reuse of existing compiler technology (fusion) for a specific purpose
%\item State of the art parallel implementation of these classes on GPUs
%\item Normalization of the grammar into efficient productions
%\item Code generator to transform a grammar into efficient code for CPU, GPU (and FPGA)
\ule
\vfill
This project has been achieved in collaboration with Manohar Jonnalagedda. I also would like to thank the LAMP team, including Eugene Burmako, Andro Stucki, Vojin Jovanovic and Tiark Rompf who provided insightful advices and suggestions. I hope you will enjoy your reading. \vspace{.3cm}\\
\textit{Thierry Coppey}

@@

\input{introduction}
\input{problems}
\input{design}
\input{implementation}
\section{Benchmarks}
XXX

@@

\section{Conclusion}

% ------------------------------------------------------------------------------------------------
\section*{Planning --- Work in progress}
\textbf{Paper introduction:} 
\begin{verbatim}
problem to solve, what exists (related work), how we solve it compared to other, how to evaluate
contributions (3): 3 tensed sentences
benchmark => prove by evaluation intro statements
\end{verbatim}

\subsubsection*{Steps}\ol
\item Re-implement vanilla combinators with lists
\item Add following problems and features:\ul
	\item Polygon triangulation (cyclic problem) (parallelogram matrix)
	\item SWat with arbitrary gap (multi-track grammar) (rectangular matrix)
	\item Zucker (rules of different complexity $O(n^2) / O(n)$)
	\ule
\item Test/proof parsers are correct --- make sure implementation is correct
\item Automate test to compare against implementation
\item Build bottom-up vanilla version\ul
	\item Separate initialization (terminals) and non-terminals (speed up?)
	\item In parallel implement the code generator for CUDA, hardcode in text function bodies for CUDA
	\ule
\item LMS implementation: from normal to Rep/Exp\ul
	\item Re-implement bottom-up vanilla into LMS
	\item Re-implement CUDA version and use functions body to infer CUDA code
	\ule
\item Benchmarks --- compare also versus other papers.
\item Write report
\ole

\subsubsection*{Roadmap}
\begin{tabular}{rl}
16.11 & Rules normalization and automatic backtracking as in 5.2 \\
	& GenScala on LMS + GenCuda + LMS CudaCompiler \\
23.11 & Problem generalization: "cyclic keyword", Zucker problem / CudaLoop optimization \\
30.11 &--- Gap due to LMS missing knowledge \\
7.12 & Benchmarking, grammar analysis \\
14.12 & First though for larger than mem \\
21.12 & Writing report \\
28.12 & --- holiday --- \\
 4.01 & --- holiday --- \\
11.01 & Writing report: implementation description and plan for future work \\
18.01 & Writing report
\end{tabular}

\subsubsection*{Todo @TCK}
\subsubsection*{Todo @Manohar}
\begin{verbatim}
Look at string templating. """xxx $var xxx"""
- Cleanup LMS CUDA generator

Separate initialization (base cases, atoms) and processing (rules, recurrences) => less cases to handle (no if)

XXX: how to encode multi-dimensional matrices
1. assume they have the same type put one after another => different dimensions ok
2. assume of same size => put into a struct
=> but using different pointers seems more reliable => completely different matrices => fixed list of matrices by dimensionality (O(1), O(n), O(n^2), ...) of structs (determined by number of indices to access object)
\end{verbatim}

\subsection*{OldPlan} \ol
\item \textbf{User facing language:} similar to \cite{adp_gpu} or \cite{adp_fusion} \href{http://hackage.haskell.org/package/ADPfusion}{\it ADP fusion}. We want to reuse the transformation mapping (problem description) $\mapsto$ (kernel implementation) for a single element.
\item \textbf{Prototyping:} understand difficulties and share common base. Prototype of \nameref{aswat} on CUDA/FPGA. Give an idea of how to implement the general case. Benchmark and compare both implementations. Aim to do as good an implementation for the specific platform (CPU/GPA) as possible.
\item \textbf{Formalize IR:} describe the intermediate representation, formalize the framework provided to the code generators (i.e. memory management, ...).
\item \textbf{Full compiler stack:} enrich the compiler stack from both top-down (translate best user-facing language parsers) and bottom-up (parametric code generators), core of the work.
\ole

\begin{verbatim}
1. optimize: terminals of bounded yield + binary decomposition if possible
2. transform into: axioms (init/fixed) + rules (iterations)
- define optimizations to IR
  - making serializable by using aggregation functions/transformations
  - avoiding the cost matrix by moving it into the wavefront
- scheduling (with memory loads for non-serializable)
- code generation for both platforms
  - possibly choice to use which platform for what part (load/compute @ cpu/gpu/fpga)

----------------------------------
Core function F:
- in: s,t strings, neighbor costs: top, left, top+left, neighbor stats: top, left, top+left
- out: backtrack information, cost(i,j), stats

we might want to pack into memory the matrix data and make threads operate on more than one cell.

Some ideas:
- define a way to pack the characters => less memory transfer (i.e. GATC=>4 letters in 1 char)
- operate on some larger word (ex 64 bits) to increase thread locality and reduce memory accesses
- write a kernel that takes stats from left and r

        stats (x)
         ||
         vv
stats -> KK -> new_stats (y')
(y)      || \
         vv  --+ backtrack info (Bxy)
     new_stats (x')

and compute the backtracking for its cell (optional as template boolean)

- multi-grain
  + within the 64b-words : multiple letters at once (thread level)
  + group threads into blocks that operate on (block level)
  + kernel that operates one after another (GPU/CPU level)
    => can we catch a stream's event at CPU level to alloc memory to get back results ?
- keep track of
  + stats O(m+n)
  + backtrack O(m*n)

- are we sure we need squares ?
  - rectangles can increase the full usage of all threads during (x-y) runs
  - assuming we put longest word vertically we play hot potato for stats
    left->right => block.x exchanges within block, last thread writes to
    global exchange memory (if short, so that it does not penalize running time)
  - we may go up to running in stripes
- if we put in diagonal-major data in the memory, why not stream all chunks like
  that using a cyclic buffer (?) => no empty slot
  ==> this might be useful for the backtrack information

- use extensively profiling : CUDA profiler

- Robust DP paper provides 2 insights: coalesced access + GPU synchronization
  => can we do a producer-consumer scenario at the block border so that we can
     execute a large diagonal horizontal/vertical swipe made of multiple blocks?
  => multiple swipes at different delays (so that we avoid syncing at every step

For stats (aggregation of the table): we need to maintain both an horizontal and vertical front
For backtracking we need to maintain the whole table
... | . |
..A | B | min/max: Min(D)=Min(Min(C),Min(B))
----+---+ sum    : Sum(D)=Sum(C)+Sum(B)-Sum(A)
..C | D | avg    : sum both #cells and values then divide at appropriate cell
----+---+
we many not need to store what's the previous cell, however, the previous
cell information (usually limited range) is much more compact than the score(32-64 bits)

----------------------------------------------------------------------------
Some GPU algorithms: http://hgpu.org/?cat=11
Translation into C++: http://bibiserv.cebitec.uni-bielefeld.de/macports/resources/download/
CUDPP libraries (but awfully big resulting binary): http://code.google.com/p/cudpp/
13 dwarfs: http://developer.amd.com/afds/assets/presentations/2155_final.pdf
http://tutorials.jenkov.com/java-reflection/fields.html
http://lampwww.epfl.ch/~michelou/scala/scala-reflection.html

XXX: use TypeClass to put a predicate on types
--------------------------------------------
def fun[T: CanTranslateToC](...)
def fun[T](implicit ev:CanTranslateToC[T])

class CanTranslateToC[T] { def translate:String }
implicit def canTranslateInt = new CanTranslateToC[Int] = { def translate = "Int" }
\end{verbatim}

>>>>

\subsection*{Abstract}
Dynamic programming is a common pattern of Computer Science used in various domains. Yet underlying matrix recurrences might be difficult to express and error prone. Additionally, domain experts might not have the skills to make an efficient parallel implementation. In this project, we present \textit{DynaProg}, a Scala DSL for dynamic programming on heterogeneous platforms which allow to write concise programs and execute them efficiently on GPUs.

Existing work is a DSL embedded in Haskell \cite{adp} with possible conversion to CUDA code \cite{adp_gpu}, a compiler for a dynamic programming external DSL into C code \cite{gapc} or ad-hoc CUDA implementations for specific problem classes \cite{swat_mega}, \cite{gpu_atlp}.
% XXX: How we compare to other, how to evaluate
% Benchmark => prove by evaluation intro statements

Our contributions are: \ul
%\item A classification of DP problems characteristics (matrix shape, dependency graph, ...)
\item A systematic approach to process data (top-down/bottom-up) and backtracking information (focus on running time and memory efficiency)
\item A language embedded in Scala (DSL) to express DP problems concisely (based on ADP)
\item Two implementations: Scala for CPU (features) and an CUDA for GPU (efficiency)
%\item Reuse of existing compiler technology (fusion) for a specific purpose
%\item State of the art parallel implementation of these classes on GPUs
%\item Normalization of the grammar into efficient productions
%\item Code generator to transform a grammar into efficient code for CPU, GPU (and FPGA)
\ule

\vfill
This project has been achieved in collaboration with Manohar Jonnalagedda. I also would like to thank the LAMP team, including Eugene Burmako, Andro Stucki, Vojin Jovanovic and Tiark Rompf who provided insightful advices and suggestions. I hope you will enjoy your reading. \vspace{.3cm}\\
\textit{Thierry Coppey}

@@

\input{introduction}
\input{problems}
\input{design}
\input{implementation}
\section{Benchmarks}
{\color{red} XXX}

@@

\begin{verbatim}
How to Encode multi-dimensional matrices efficiently
1. assume they have the same type put one after another => different dimensions ok
2. assume of same size => put into a struct
=> but using different pointers seems more reliable => completely different matrices => fixed list of matrices by dimensionality (O(1), O(n), O(n^2), ...) of structs (determined by number of indices to access object)

Possible extensions / optimizations:
- making serializable by using aggregation functions/transformations
- avoiding the cost matrix by moving data in the wavefront
- Scheduling for large problems (with memory loads for non-serializable)
- add FPGA target platform
- pack the data => less memory transfer (i.e. GATC=>4 letters in 1 char)
- operate on some larger word (ex 64 bits) to increase thread locality and reduce memory accesses

Core function F:
- in: s,t strings, neighbor costs: top, left, top+left, neighbor stats: top, left, top+left
- out: backtrack information, cost(i,j), stats
        stats (x)
         ||
         vv
stats -> KK -> new_stats (y')
(y)      || \
         vv  --+ backtrack info (Bxy)
     new_stats (x')

For stats (aggregation of the table): we need to maintain both an horizontal and vertical front
For backtracking we need to maintain the whole table
... | . |
..A | B | min/max: Min(D)=Min(Min(C),Min(B))
----+---+ sum    : Sum(D)=Sum(C)+Sum(B)-Sum(A)
..C | D | avg    : sum both #cells and values then divide at appropriate cell
----+---+
we many not need to store what's the previous cell, however, the previous
cell information (usually limited range) is much more compact than the score(32-64 bits)
\end{verbatim}

\section{Conclusion}
{\color{red} XXX}

% ------------------------------------------------------------------------------------------------
\section*{Planning}
\textbf{Paper introduction}\ul
\item Problem to solve, what exists (related work), how we compare to other, how to evaluate
\item Contributions (3): 3 tensed sentences
\item benchmark => prove by evaluation intro statements
\ule

\subsubsection*{Roadmap}
\begin{tabular}{rl}
16.11 & Rules normalization and automatic backtracking \\
	& GenScala on LMS + GenCuda + LMS CudaCompiler \\
23.11 & Problem generalization: "cyclic keyword", Zucker problem / CudaLoop optimization \\
30.11 &--- Gap due to LMS missing knowledge \\
7.12 & Benchmarking, grammar analysis \\
14.12 & First thoughts for larger than device memory \\
21.12 & Writing report \\
28.12 & --- holiday --- \\
 4.01 & --- holiday --- \\
11.01 & Writing report: implementation description and plan for future work \\
18.01 & Writing report
\end{tabular}

\subsubsection*{Todo @TCK}\ul
\item Test/proof parsers are correct --- make sure implementation is correct
\item Automate test to compare against implementation
\item Benchmarks -- use extensively CUDA profiler
\item Write report
\item Port LibRNA for CUDA?
\ule

\subsubsection*{Todo @Manohar}\ul
\item Integrate LMS code generation into v4.
\item Fix Zuker coefficients
\ule

%Some GPU algorithms: http://hgpu.org/?cat=11
%Translation into C++: http://bibiserv.cebitec.uni-bielefeld.de/macports/resources/download/
%CUDPP libraries (but awfully big resulting binary): http://code.google.com/p/cudpp/
%13 dwarfs: http://developer.amd.com/afds/assets/presentations/2155_final.pdf
%http://tutorials.jenkov.com/java-reflection/fields.html
%http://lampwww.epfl.ch/~michelou/scala/scala-reflection.html
%
%Hint: use TypeClass to put a predicate on types
%  def fun[T: CanTranslateToC](...)
%  def fun[T](implicit ev:CanTranslateToC[T])
%  class CanTranslateToC[T] { def translate:String }
%  implicit def canTranslateInt = new CanTranslateToC[Int] = { def translate = "Int" }
